---
title: "Assignment 3 - CRD"
author: "Polycarpe Mizero"  
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
    number-sections: false
    theme: cerulean
---

# Introduction  
This assignment will ask you to conduct a full CRD analysis, from data import through publication ready plot.  

For that, you will first need to watch (and code along with) the following CRD videos:

- CRD lecture: https://youtu.be/r5vsakgqMXc?si=8udshzlpZ-8td1f4&t=4941 (starts at 1:22:21, watch till end)

- CRD lecture pt 2, lab pt 1: https://youtu.be/GpfHxh1eK5U?si=nTle5YcB8JQklja2 
(watch whole video)  

- CRD lab pt 2: https://youtu.be/YQ1bK6uV194?si=LnDzRT69OsiMLc-r
(wathc the whole video)  

**AFTER watching the videos and developing the code in class**, then start working on this HW assignment.  

The different sections representing what we did in class were left in the script to provide you a scaffold to start from.  

# Instructions  
  - Move this script into the `code` subfolder of your `03_crd` project folder  
  - Move the file `wheat_nk_bamyan.csv` into the `data` folder of your `03_crd` project folder  
  - On this script, fill your name under the `title` section of the YAML  

# a) Setup  
Here is where we load the packages we will use.  
Make sure to load each package in the chunk below.  
```{r setup, message=FALSE, warning=FALSE}

library(tidyverse) # for data wrangling and plotting
library(car) # for Anova function
library(broom) # for model residuals extraction
library(emmeans) # for model mean extraction
library(multcomp) # for pairwise comparison letter display

```

Now, import the data set `wheat_nk_bamyan.csv` in the chunk below.  
```{r data import}
crd_df <- read_csv("../data/wheat_nk_bamyan.csv")

crd_df
printrows <- crd_df[1:6,]
printrows
```

The **objective of this study** was to assess the effect of different N and K rates on wheat yield.  

For that, a study was conducted using a completely randomized design as the experimental design with four replicates. The treatment design was a 3 N x 3 K rate factorial, for a total of 9 treatments.  


# b) EDA tables  
Use this space to conduct exploratory data analysis of the imported data, in the form of tables.  
```{r summary}
summary(crd_df)

```
```{r glimpse}
glimpse(crd_df)
```


# c) Wrangling  
Here, make sure to wrangle your data set with all needed steps. Refer back to class code if you need a refresher on what was needed.  

```{r wrangled data}
crd_dfw <- crd_df %>%
  mutate(rep = factor(rep),
         nrate_kgha = factor(nrate_kgha),
         krate_kgha = factor(krate_kgha)
         ) %>%
  mutate(trtname = paste0(nrate_kgha, "+", krate_kgha))

crd_dfw
printrows <- crd_dfw[1:6,]
printrows
```
```{r wrangled data summary}
summary(crd_dfw)
```
Number of replication: 4
Number of treatment: 3 N rates X 3 k rates = 9
Number of observations: 4X 9 = 36
Yield: 812 to 7232 kg/ha

# d) EDA plots  
Now, explore the wrangled data using plots.  
Make sure to explore plots showing the main effects and the interaction.  

## Question 1) Below each plot of this section, write a sentence explaining your visual interpretation of the plot.  

```{r n_rate_kgha boxplot}
ggplot(crd_dfw, aes(x = nrate_kgha,
                    y = yield_kgha,
                    color = nrate_kgha)) +
  geom_boxplot()+
  geom_jitter() +
  theme(legend.position = "none")
```
This box plot shows the relationship between nitrogen rate (nrate_kgha) and wheat yield (yield_kgha), demonstrating that as nitrogen rate increases from 0 to 100 to 200 kg/ha, yield increases, implying that higher nitrogen rates tend improve yield. There is also some overlap between the yield ranges at different nitrogen rates, indicating that nitrogen is not the only factor influencing yield.

```{r k_rate_kgha boxplot}
ggplot(crd_dfw, aes(x = krate_kgha, 
                    y = yield_kgha,
                    color = krate_kgha)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")
```
The box plot above shows the relationship between different potassium rates (krate_kgha) and yield (yield_kgha), displaying a yield increases as potassium application increases from 0 to 30 to 60 kg/ha. The highest yield is observed at krate_kgha = 60 kg/ha, suggesting that higher potassium levels improve yield.

```{r nk interaction boxplot}
ggplot(crd_dfw, aes(x = nrate_kgha,
                    y = yield_kgha,
                    color = nrate_kgha )) +
  geom_boxplot() +
  geom_jitter() +
  facet_grid(.~krate_kgha)
```
The box plot above shows the interaction effect of different nitrogen rates inside potassium levels. There is a yield increase with both nitrogen and potassium.At higher rates of potassium (30 and 60 kg/ha) there is a significant increase in yield, even at lower nitrogen levels, suggesting that potassium may be more effective when nitrogen is not limiting. From the graph, the optimal yield can achieved at high nitrogen (100–200 kg/ha) and high potassium (60 kg/ha).

```{r}
ggplot(crd_dfw, aes(x = krate_kgha,
                    y = yield_kgha,
                    color = krate_kgha )) +
  geom_boxplot() +
  geom_jitter() +
  facet_grid(.~nrate_kgha)
```
Interaction effect of different rates of potassium inside different levels on nitrogen. 
At higher nitrogen levels (100 and 200 kg/ha), there is a stronger effect of potassium on yield, but their interaction is not that significant.
From the graph, at higher rates of nitrogen (100 and 200 kg/ha), adding potassium significantly boosts yield.

# e) Statistical model  
In this section, run the appropriate model.  
Remember to set the contrast type to sum-to-zero (the default in R is set-to-zero, which is not what we want).  

Also, remember to specify all model terms needed for the treatment design.  

```{r model}
# Changing to sum-to-zero contrast
options(contrasts = c("contr.sum", "contr.poly"))

# Model fitting
crd_mod <- lm(yield_kgha ~ nrate_kgha + krate_kgha +
                nrate_kgha:krate_kgha,
              data = crd_dfw
                )

crd_dfw

# Summary
summary(crd_mod)
```
Model contrast set sum-to-zero

## Model specification tips

# f) ANOVA table  
Use this section to check the ANOVA table, ensuring you use type 3 sum-of-squares.

```{r ANOVA}
Anova(crd_mod, type = 3)
```
This ANOVA table evaluates the effects of nitrogen rate (nrate_kgha), potassium rate (krate_kgha), and their interaction (nrate_kgha:krate_kgha) on yield (yield_kgha).  

Starting with the interaction effect (nrate_kgha:krate_kgha), with F value = 0.7954 and p-value = 0.53857 (not significant), this indicates that the interaction between nitrogen and potassium rates is not statistically significant. This suggests that the effect of nitrogen on yield does not depend on the level of potassium applied (or vice versa).  

Considering the main effect of potassium rate (krate_kgha) with F value = 19.44, and p-value = 5.89e-06 (highly significant). This indicate that potassium rate has a strong and highly significant effect on yield. Meaning that varying potassium levels significantly impact crop yield.  

When it comes to the main effect of noitrogen rate (nrate_kgha) with F value = 6.84, and p-value = 0.00394 (Significant).This suggest that the nitrogen rate significantly affects yield, indicating that different nitrogen application rates lead to variations in yield.  

The intercept has F value = 416.45, and p-value: < 2.2e-16 (highly significant). The intercept is significantly different from zero, meaning there is a strong baseline yield when all factors are at their reference levels.


## Question 2) Based on the ANOVA table above, for which effect(s) in the ANOVA should you extract means and perform pair-wise comparisons? Explain why.  

Based on the ANOVA table, we should extra means of the main effect of nitrogen rates (nrate_kgha) and the main effect of potassium rates (krate_kgha), and perform their pairwise comparison.  
This is because both nitrogen rate and potassium rate significantly affect yield, with potassium having a stronger effect (F = 19.44 vs. F = 6.84).  
In addition, the interaction between nitrogen and potassium is not significant, indicating that their effects on yield are independent.

# g) Linear model assumptions  
Use this section to first compute your residuals, including studentdized residuals.

```{r crd_resid}
crd_resid <- augment(crd_mod) %>%
  mutate(.studresid = rstudent(crd_mod))

crd_resid
```


## Question 3) What are the linear model assumptions that we need to check for? Create subsections (using ##) below for each of these assumptions, and write code to produce plots to check for these assumptions. For each assumption, write down your interpretation of the produced plots, and your thoughts if the assumptions hold or not.  

Here below are the linear model assumptions that we need to check for:

- Residual independence (no pattern)  
- Residual variance homogeneity (homoscedasticity)  
- Residual normality   
- Outlier detection (< -3 or > 3) 

## Residual independence (no pattern)

```{r}
ggplot(crd_resid, aes(x = .fitted,
                      y = .studresid)) +
  geom_point(shape = 21,
             fill = "green",
             size = 3,
             alpha = .7
             ) +
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth()
```
The graph shows that residuals are randomly distributed and do not follow any pattern.  
Therefore, the residual independence assumption is verified.

## Residual variance homogeneity (homoscedasticity) 

```{r}
ggplot(crd_resid, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "green", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  theme_bw()
```
Residual are randomly scattered around zero (.studresid = 0), showing no pattern, and therefore satisfying the homoscedasticity.


## Residual normality (discuss this!)


```{r}
ggplot(crd_resid, aes(sample = .studresid)) +
  stat_qq() +
  stat_qq_line()

```
Residuals closely follow the diagonal black line indicating that theresiduals are normally distributed.  
there is a slight deviation at the upper and lower tails of the distribution suggest some outliers having a minor non-normality in extreme values, but this should not be a major concern.  
Since most points are aligned with the line, and the residuals appear to be centered around zero and normally distributed, the assumption for of residual normality is largely satisfied.


```{r density plot}
ggplot(crd_resid, aes(x = .studresid)) +
  geom_density() +
  scale_x_continuous(breaks = c(-3,0,3), limits = c(-3,3))

```
This plot represents the density distribution of studentized residuals. 
The peak of the density curve is around 0 which is a good indication that the residuals have a mean close to zero.  
The curve shows a slight second peak around 1 which may suggest some minor deviations, but having a bigger peak around 0 satisfies the assumption that resiaduals are normally distributed with a mean of zero.

## Residual outlier detection (< -3 or > 3)


```{r}
ggplot(crd_resid, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 21,
             fill = "green", 
             size = 3,
             alpha = .7)+
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  theme_bw()
```
All residuals are within the [-3, 3] interval, so nothing to worry here.  
Now that model assumptions have been checked and met, we can proceed to using the model for inference. 

# h) Model means  
Use this section to extract model means according to your decision of the ANOVA table effects.  

```{r means for n rate, message=FALSE, warning=FALSE}
crd_means_n <- emmeans(crd_mod, ~nrate_kgha)

crd_means_n
```


```{r means for k rate, message=FALSE, warning=FALSE}
crd_means_k <- emmeans(crd_mod, ~krate_kgha)

crd_means_k
```


# i) Pairwise comparisons  
Use this section to perform pair-wise comparison of the means extracted above.  

Make sure to reverse the order from high to low, use letters instead of numbers, and no adjustment. 


```{r pwc for n rate}
crd_cld_n <- cld(crd_means_n, 
                   reversed=T, 
                   adjust="none",
               Letters= letters)

crd_cld_n
```

```{r pwc for k rate}
crd_cld_k <- cld(crd_means_k, 
                   reversed=T, 
                   adjust="none",
               Letters=letters) 

crd_cld_k
```


# g) Final plot(s)  
Produce final plot(s) to show your results.  
Make sure to include both **raw data** and **statistical model summary (i.e., letter separation)**.  

Be mindful of your plot to ensure you are creating it in an intuitive way for your audience in relation to interpreting the letter separation structure you used.  

In class, we did not have time to make it publication-ready (e.g., a plot looking professional and ready to be included in a manuscript).

Here, make sure you deliver final plot(s) that is/are publication ready.  

Make sure to provide an interpretation for the plot(s) created, including which treatment levels were significantly higher/lower.

```{r  final plot for n rate}
ggplot() +
geom_boxplot(data = crd_dfw,
             aes(x = nrate_kgha, 
                 y = yield_kgha,
                 fill = nrate_kgha
                 )) +
geom_jitter(data = crd_dfw, aes(x = nrate_kgha, y= yield_kgha)) +
geom_label(data = crd_cld_n,
           aes(x = nrate_kgha,
               y = emmean,
               label = .group)) + 
labs(x= "N rate (kg/ha)", y = "Yield (kg/ha)", title = "Effect of N rate on wheat yield") +
  theme_linedraw()
  

```
The letters inside the box plots indicate statistical grouping based on different rates of nitrigen on wheat yield.
The letters indicate significant differences in yield when n_rate_kgha = 0 kg/ha, while when n_rate_kgha = 100 and 200 kg/ha indicate no significant differences.  
Taking an example from where n = 100 kg/ha, the yield is slightly similar with the yield at n= 200 kg/ha.


```{r final plot for k rate}

ggplot() +
geom_boxplot(data = crd_dfw,
             aes(x = krate_kgha, 
                 y = yield_kgha,
                 fill = krate_kgha
                 )) +
geom_jitter(data = crd_dfw, aes(x = krate_kgha, y= yield_kgha)) +
geom_label(data = crd_cld_k,
           aes(x = krate_kgha,
               y = emmean,
               label = .group)) + 
labs(x= "K rate (kg/ha)", y = "Yield (kg/ha)", title = "Effect of K rate on wheat yield") +
  theme_linedraw()

```
This boxplot show the effects of differnt rates of potassium (k) on wheat yield (kg/ha).  

Yield increases as k application rate increases, but at higher potassium levels (k_rate_kgha = 30 and 60 kg/ha), the yield is almost similar.   
Outliers (dots) are observed in some cases especially at higher potassium levels.



# Submitting your work  
Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE` (I'll demo this in a moment).  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one (I'lld demo this in a moment).  

After rendering, an .html file will be created on your `code` folder.  

Rename this file to `Assignment-03-crd-LASTNAME.html`.    

For ex., mine would be `Assignment-03-crd-Bastos.html`.

Submit your work by uploading the **html** file to **eLC Assigment #3 - CRD** by Feb 3rd 11:59 pm.  
  




